Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-13 Manuscript/sec2-review-of-graph-neural-network.tex
Update the ChebNet complexity analysis.

Different GNNs adopt different kinds of graph neurons and have different definitions of the three functions.
$\phi$ and $\Sigma$ are the \emph{edge computation} functions.
They are conducted over every edge in $\mathcal{G}$.
$\gamma$ is the \emph{vertex computation} function.
It is conducted over every vertex in $\mathcal{G}$.
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge functions and the vertex functions of typical GNNs, respectively.

\begin{table}
	\hspace{-10em}
    \begin{footnotesize}
        \begin{tabular}{ccp{8em}p{22em}r}
		\toprule
		GNN &
		Type &
		$\Sigma$ &
		$\phi$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i, , k}^l = T_k(\widetilde{L} )_{j, i} \boldsymbol{h}_j^l$ &
		$O(K * d_{in})$ \\
		\textbf{GCN} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = e_{j, i} \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		AGCN &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \tilde{e}_{j, i}^l \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		GraphSAGE &
		Non-spectral &
		mean/LSTM &
		$\boldsymbol{m}_{j, i}^l =  \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		Neural FPs &
		Non-spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		SSE &
		Recurrent &
		sum &
		$\boldsymbol{m}_{j, i}^l = [\boldsymbol{h}_i^{l} \parallel \boldsymbol{h}_j^l]$ &
		$O(1)$ \\
		\textbf{GGNN} &
		Gated &
		sum &
		$\boldsymbol{m}_{j, i} = \textcolor{blue}{\boldsymbol{W}^l} \boldsymbol{h}_j^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		Attention &
		sum  &
		\begin{scriptsize}
			$\begin{aligned}[t]
			& \alpha_{j, i}^k = \frac {\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l] ))} {\sum_{k \in \mathcal{N}(i)}\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_k^l] ))} \\
			&  \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})
			\end{aligned}$
		\end{scriptsize} &
		$O(K * d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		Attention &
		sum,max,mean &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\alpha_{j, i}^k = \frac {\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l, k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}} \boldsymbol{h}_i^l] )} {\sum_{k \in \mathcal{N}(j)}\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}}  \boldsymbol{h}_k^l] )} \\
			&\boldsymbol{m}_{j, i, 1}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}_v} \boldsymbol{h}_j^{l}) \\
			&\boldsymbol{m}_{j, i, 2}^l = \textcolor{blue}{\boldsymbol{W}^l_m} \boldsymbol{h}_j^{l} \\
			&\boldsymbol{m}_{j, i, 3}^l = \boldsymbol{h}_j^l
			\end{aligned}$
		\end{scriptsize} &
		$O(max(d_a, d_m) * K * d_{in})$ \\
		\bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their edge computation functions.$d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively. Blue variables are model parameters to learn.}
	\label{tab:gnn_overview_edge}
\end{table}

\begin{table}
	\hspace{-3em}
    \begin{footnotesize}
        \begin{tabular}{cp{20em}r}
		\toprule
		GNN &
		$\gamma$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		$\boldsymbol{h}_i^{l+1} = \sum_{k=0}^K \textcolor{blue}{\boldsymbol{W}}^{l,k}  \boldsymbol{s}_{i, k}^{l} $ &
		$O(K * d_{in} * d_{out})$ \\
		\textbf{GCN} &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		AGCN &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE &	
		$\boldsymbol{h}_i^{l+1} =   \delta(\textcolor{blue}{\boldsymbol{W}}^l  [\boldsymbol{s}_i^{l} \parallel \boldsymbol{h}_i^l])$ &
		$O(d_{in} * d_{out})$ \\
		Neural FPs &
		$\boldsymbol{h}_i^{l+1} = \delta(\boldsymbol{h}_i^l + \textcolor{blue}{\boldsymbol{W}}^{\boldsymbol{N}_i}  \boldsymbol{s}_i^{l})$ &
		$O(d_{in} * d_{out})$ \\
		SSE &
		$\boldsymbol{h}_i^{l+1} = (1 - \alpha)  \boldsymbol{h}_i^l +\alpha    \delta(\textcolor{blue}{\boldsymbol{W}}^l_1 \delta(\textcolor{blue}{\boldsymbol{W}}^l_2), \boldsymbol{s}_i^l)$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GGNN} &
		\begin{scriptsize}
		$\begin{aligned}[t]
			& \textcolor{blue}{\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^z \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^{sz} + \textcolor{blue}{\boldsymbol{U}}^z \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hz}) \\ 
			&\boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^r \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}}^{sr} +\textcolor{blue}{\boldsymbol{U}}^r \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hr}) \\ 
			&\boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}}^l \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^s + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^h))) \\ 
			&\boldsymbol{h}_i^{l+1} = (1 - \textcolor{blue}{\boldsymbol{z}}_i^l) \odot \boldsymbol{h}_i^l +  \textcolor{blue}{\boldsymbol{z}}_i^l \odot \boldsymbol{h}_i^{l+1}
		\end{aligned}$
		\end{scriptsize}
		&
		$O(max(d_{in}, d_{out}) * d_{out})$ \\
		\textbf{GAT}&
		$\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(1)$ \\
		\textbf{GaAN} &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\boldsymbol{g}_i = \textcolor{blue}{\boldsymbol{W}}^l_g  [\boldsymbol{h}_i^{l} \parallel \boldsymbol{s}_{i, 2}^l \parallel \boldsymbol{s}_{i, 3}^l]  \\
			&\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l_o [\boldsymbol{h}_i^l \parallel (\boldsymbol{g}_{i} \odot \boldsymbol{s}_{i, 3}^l) ]
			\end{aligned}$ 
		\end{scriptsize}&
		$O(max(d_{in} + K * d_v, 2 * d_{in} + d_m) d_{out})$ \\ \bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their vertex computation functions. $d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively. Blue variables are model parameters to learn.}
	\label{tab:gnn_overview_vertex}
\end{table}

>>>>

Different GNNs adopt different kinds of graph neurons and have different definitions of the three functions.
$\phi$ and $\Sigma$ are the \emph{edge computation} functions.
They are conducted over every edge in $\mathcal{G}$.
$\gamma$ is the \emph{vertex computation} function.
It is conducted over every vertex in $\mathcal{G}$.
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge functions and the vertex functions of typical GNNs, respectively.
For ChebNet, we report its GNN layer in the tables \footnote{A layer of ChebNet consists of $K$ GNN sub-layers and a summation layer, i.e., $\boldsymbol{H}^{l+1} = \sum_{k=1}^K{\boldsymbol{Z}^{(k)} \textcolor{blue}{\boldsymbol{W}^{(k)}}}$ with GNN layers $\boldsymbol{Z}^{(1)}=\boldsymbol{H}$, $\boldsymbol{Z}^{(2)}=\hat{\boldsymbol{L}}\boldsymbol{H}$, and $\boldsymbol{Z}^{(k)}=2\hat{\boldsymbol{L}}\boldsymbol{Z}^{(k-1)} - \boldsymbol{Z}^{(k-2)}$. $\boldsymbol{H}^l$ is the matrix of output hidden feature vectors of the layer $l-1$. In the table, we report the GNN sub-layer of ChebNet that calculates $\boldsymbol{Z}^{(k)}$}.

\begin{table}
	\hspace{-10em}
    \begin{footnotesize}
        \begin{tabular}{ccp{8em}p{22em}r}
		\toprule
		GNN &
		Type &
		$\Sigma$ &
		$\phi$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^{(k)} = e_{j, i}\boldsymbol{z}_j^{(k-1)}$ &
		$O(d_{in})$ \\
		\textbf{GCN} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = e_{j, i} \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		AGCN &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \tilde{e}_{j, i}^l \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		GraphSAGE &
		Non-spectral &
		mean/LSTM &
		$\boldsymbol{m}_{j, i}^l =  \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		Neural FPs &
		Non-spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		SSE &
		Recurrent &
		sum &
		$\boldsymbol{m}_{j, i}^l = [\boldsymbol{h}_i^{l} \parallel \boldsymbol{h}_j^l]$ &
		$O(1)$ \\
		\textbf{GGNN} &
		Gated &
		sum &
		$\boldsymbol{m}_{j, i} = \textcolor{blue}{\boldsymbol{W}^l} \boldsymbol{h}_j^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		Attention &
		sum  &
		\begin{scriptsize}
			$\begin{aligned}[t]
			& \alpha_{j, i}^k = \frac {\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l] ))} {\sum_{k \in \mathcal{N}(i)}\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_k^l] ))} \\
			&  \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})
			\end{aligned}$
		\end{scriptsize} &
		$O(K * d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		Attention &
		sum,max,mean &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\alpha_{j, i}^k = \frac {\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l, k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}} \boldsymbol{h}_i^l] )} {\sum_{k \in \mathcal{N}(j)}\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}}  \boldsymbol{h}_k^l] )} \\
			&\boldsymbol{m}_{j, i, 1}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}_v} \boldsymbol{h}_j^{l}) \\
			&\boldsymbol{m}_{j, i, 2}^l = \textcolor{blue}{\boldsymbol{W}^l_m} \boldsymbol{h}_j^{l} \\
			&\boldsymbol{m}_{j, i, 3}^l = \boldsymbol{h}_j^l
			\end{aligned}$
		\end{scriptsize} &
		$O(max(d_a, d_m) * K * d_{in})$ \\
		\bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their edge computation functions.$d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively.
		Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.}
	\label{tab:gnn_overview_edge}
\end{table}

\begin{table}
	\hspace{-3em}
    \begin{footnotesize}
        \begin{tabular}{cp{20em}r}
		\toprule
		GNN &
		$\gamma$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		$\boldsymbol{z}_i^{(k)} = 2\boldsymbol{s}^{(k)}_{i} - \boldsymbol{z}_i^{(k-2)}$&
		$O(d_{in})$ \\
		\textbf{GCN} &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		AGCN &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE &	
		$\boldsymbol{h}_i^{l+1} =   \delta(\textcolor{blue}{\boldsymbol{W}}^l  [\boldsymbol{s}_i^{l} \parallel \boldsymbol{h}_i^l])$ &
		$O(d_{in} * d_{out})$ \\
		Neural FPs &
		$\boldsymbol{h}_i^{l+1} = \delta(\boldsymbol{h}_i^l + \textcolor{blue}{\boldsymbol{W}}^{\boldsymbol{N}_i}  \boldsymbol{s}_i^{l})$ &
		$O(d_{in} * d_{out})$ \\
		SSE &
		$\boldsymbol{h}_i^{l+1} = (1 - \alpha)  \boldsymbol{h}_i^l +\alpha    \delta(\textcolor{blue}{\boldsymbol{W}}^l_1 \delta(\textcolor{blue}{\boldsymbol{W}}^l_2), \boldsymbol{s}_i^l)$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GGNN} &
		\begin{scriptsize}
		$\begin{aligned}[t]
			& \textcolor{blue}{\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^z \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^{sz} + \textcolor{blue}{\boldsymbol{U}}^z \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hz}) \\ 
			&\boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^r \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}}^{sr} +\textcolor{blue}{\boldsymbol{U}}^r \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hr}) \\ 
			&\boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}}^l \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^s + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^h))) \\ 
			&\boldsymbol{h}_i^{l+1} = (1 - \textcolor{blue}{\boldsymbol{z}}_i^l) \odot \boldsymbol{h}_i^l +  \textcolor{blue}{\boldsymbol{z}}_i^l \odot \boldsymbol{h}_i^{l+1}
		\end{aligned}$
		\end{scriptsize}
		&
		$O(max(d_{in}, d_{out}) * d_{out})$ \\
		\textbf{GAT}&
		$\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(1)$ \\
		\textbf{GaAN} &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\boldsymbol{g}_i = \textcolor{blue}{\boldsymbol{W}}^l_g  [\boldsymbol{h}_i^{l} \parallel \boldsymbol{s}_{i, 2}^l \parallel \boldsymbol{s}_{i, 3}^l]  \\
			&\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l_o [\boldsymbol{h}_i^l \parallel (\boldsymbol{g}_{i} \odot \boldsymbol{s}_{i, 3}^l) ]
			\end{aligned}$ 
		\end{scriptsize}&
		$O(max(d_{in} + K * d_v, 2 * d_{in} + d_m) d_{out})$ \\ \bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their vertex computation functions. $d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively. Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.}
	\label{tab:gnn_overview_vertex}
\end{table}
